SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-11-19 method.tex
Mostly editing for English, redundancy, some technical changes, and redundancy.

\section{METHODOLOGY}
\label{Method}

\subsection{Performance Model}
\label{Method-Model}
For testing and comparing NATCH, we used ChampSim~\cite{Champsim}
simulator.  ChampSim is an enhanced version of the framework that was
used for the 2nd Data Prefetch Championship (DPC-2)~\cite{DPC_2}. We
model a 1-core / 4-core / 8-core Out of Order machine. The
details of the configuration parameters is summarized in Table
\ref{tab:Sim_params}.

@@

The block size was fixed 64KB. Prefetching was only triggered at L2
Cache demand accesses but could be directed to L2 Cache or Last Level
Cache. No L1 Data level prefetching was done. LRU replacement policy
was used on all levels of cache hierarchies. Branch prediction was
done using the perceptron branch predictor~\cite{Perc_Branch}.  Page
size was kept as 4KB.  Champsim operates all the prefetchers strictly
in the physical address space.

@@

\subsection{Workloads}
\label{Method-Workloads}
This is the first time that SPEC 2017 benchmark suite~\cite{SPEC2017}
has been used to characterize and measure the prefetch performance.
We use all the 20 workloads available in the SPEC 2017 Suite.  Using
SimPoint~\cite{SimPoint} methodology, we identified 95 different
program segments of 1 Billion instructions each.

\textit{Single-core performance:} For single-core simulations, we use
200 Million instructions to warm-up the microarchitectural structures
and next 1 Billion instructions to do detailed simulations and collect
run-time statistics. We report the IPC speed over
baseline \textit{i.e.}, no prefetching.
The final number reported is the geometric mean of the speed-up
attained on individual traces.

\textit{Multi-core performance:} For multi-application workloads, we
generate 100 random mixes and another 100 mixes from the
memory intensive subset of SPEC 2017.  For 4-core workload, 200
Million instructions are used for warm-up and additional 1 Billion
instruction simulated for collecting statistics.  Each CPU keeps
executing its workload till the last CPU completes 1 Billion
instructions after warm-up.  For collecting IPC and other data, only
the first billion instructions are considered as the ``region of
interest''.

@@

We repeat the same process for 8-core workloads, correspondingly with 16MB LLC.
Only exception is that 20 Million warm-up instructions and 100 Million 
full instructions were executed.  
This was done so as to keep the simulation run-time
within reasonable limits as a single 8-core mix was taking up to 3
days to simulate 1 Billion instructions.

\textit{Cross Validation:} We cross-validated our NATCH model using SPEC
2006~\cite{SPEC2006} and CloudSuite~\cite{CloudSuite} benchmarks.  
For single core
SPEC 2006, we developed 94 simpoints spread across all the 29
applications. For multi-core, 
we followed the same methodology as SPEC 2017. 
For CloudSuite, we used the traces made available for
the 2nd Cache Replacement Competition (CRC-2)~\cite{CRC_2}.  The traces
include 4 4-core applications with 6 distinct phases per
application.

@@

\subsection{Preferchers Simulated}
\label{Method-Prefetchers}
We compared NATCH against three of the latest state of the art
hardware-only prefetchers: Best Offset Prefetcher (BOP), DRAM Aware - Access Map
Pattern Matching (DA-AMPM)~\cite{DA_AMPM} and Signature Path Prefecher
(SPP).  BOP was the winner of 2nd Data Prefetching Championship.
DA-AMPM is the enhanced version of AMPM, modified to account for DRAM
row buffer locality.  SPP has been shown to outperform BOP on SPEC
2006 traces.  For each of these, we compare their speedups taking no
prefetching as the baseline.

\subsection{Developing Features for NATCH}
\label{Method-Features}
This section describes the intuition and analysis that went behind
finalizing the features.  As noted earlier, we developed a set of 9
features which allow the perceptron throttler to correlate prefetching
decision with the program context.  To study the correlation across
each feature, we study statistically the perceptron weights and try to
interpret their distribution.

\textbf{GLOBAL PEARSON'S CORRELATION}\newline For this experiment, we
create a dump of the state of perceptron weights at the end of all
trace execution.  The weights obtained from running all the SPEC 2017
traces are concatenated together.  Since the weights are collected at
the end of individual trace execution, the perceptron weights have
attained a relatively stable value by now.  Hence, this dump
represents a `snap' of the trained perceptron weights across all the
SPEC 2017 traces.  The intuition here is that the feature with bulk of
the perceptron weights concentrated around 0 or small magnitude
numbers show a weak correlation with the prefetching outcome.  On the
other hand, features with most of the weights saturated around highest
value (+15) show a high positive correlation and the features with
weights close to the lowest value (-16) show a strong negative
correlation.

@@

\textbf{PER TRACE CORRELATION} \newline Another important way to look
at the perceptron features is to see how much their contribution
varies across the traces.  Here we give special attention to features
with low P-values in the previous experiment.  Figure XX shows the
variation of P-values three features : \_\_, \_\_ and \_\_; across all
the SPEC 2017 traces.  For simplicity, the traces are arranged in an
increasing order of contribution made by the feature and the trace
names have been omitted.  It can be seen that even features with a low
overall correlation provide useful correlation (magnitude > 0.4) for
XX out of the 95 simpoints developed for SPEC 2017.

\textbf{TRIMMING FEATURES USING CORRELATION} \newline Besides
providing interesting insights into prefetching behaviour, P-value can
also be used for feature selection and prefetcher tuning.

Here, we introduce the concept of cross-correlation across features.
Just as we examined correlation of each feature with the final
outcome, we can also study correlation between the features.  We used
the above methodology to eliminate features which provide a little or
no information that has already been captured in other features.

As a part of feature selection, we initially came up a mix of 23
features- primary and composite.  By studying cross correlation of
each of these features against others in a 23x23 matrix, we identified
pairs of features with correlation factor > 0.9 in magnitude and
carefully eliminated redundant features.  Using this approach, we
managed to reduce the feature count to 9.  Thus, in the final
implementation of NATCH, no two features have a high correlation
between them.  This way we can be sure that each feature is bringing
in contribution that cannot be captured using other features.

@@

\subsection{Overhead for NATCH}
\label{Method-Overheads}

@@

In this section, we analyze the hardware overhead required to
implement NATCH.  The Prefetch Filter was enhanced to accommodate
storing of meta-data for perceptron training.  Table
\ref{tab:PF_metadata} depicts the meta-data stored for each entry in
the Prefetch Filter.  Table \ref{tab:NATCH_overhead} shows the total
storage overhead of NATCH implementation.  The hardware budget for
2nd Data Prefetching championship was 32 KB.  Keeping that in mind 
the considerable speedup NATCH obtained over the winner, the extra hardware
budget can be accounted for.  The extra hardware also makes the
overall scheme more scalable than SPP.  In the original SPP paper, it
was demonstrated that adding extra hardware brings little advantage in
terms of performance gain.  The newly added perceptron tables can be
scaled to increase / decrease features depending on the permitted
budget.

>>>>

\section{Methodology}
\label{Method}

\subsection{Performance Model}
\label{Method-Model}
For testing and comparing PPF, we used the ChampSim~\cite{Champsim} simulator.
ChampSim is an enhanced version of the framework that was used for the 2nd
Data Prefetch Championship (DPC-2)~\cite{DPC_2}. We model 1-core, 4-core, and
8-core out-of-order machines. The details of the configuration parameters are
summarized in Table~\ref{tab:Sim_params}.

@@

% djimenez: changing all the "was" to "is." when describing our work, use
% present tense

The block size is fixed at 64 bytes. Prefetching is only triggered at an L2 cache
demand accesses but could be directed to the L2 or last-level cache. No L1
data level prefetching is done. The LRU replacement policy is used on all levels
of cache hierarchies. Branch prediction is done using the perceptron branch
predictor~\cite{Perc_Branch}.  ThE page size is 4KB.  Champsim operates
all the prefetchers strictly in the physical address space.

@@

\subsection{Workloads}
\label{Method-Workloads}
% djimenez: don't say this is the first time SPEC 2017 has been used this way.
% i don't know if that's true, and in any event the value is small.

%This is the first time that SPEC 2017 benchmark suite~\cite{SPEC2017}
%has been used to characterize and measure the prefetch performance.
We use the that SPEC 2017 benchmark suite~\cite{SPEC2017}.  We use all the 20
workloads available in the SPEC 2017 suite.  Using the SimPoint~\cite{SimPoint}
methodology, we identified 95 different program segments of 1 Billion
instructions each.

\textit{Single-core performance:} For single-core simulations, we use 200
million instructions to warm-up the microarchitectural structures and the next
one billion instructions to do detailed simulations and collect run-time
statistics. We report the IPC speedup over the baseline of no prefetching.
The final number reported is the geometric mean of the speedup achieved on
individual traces.

\textit{Multi-core performance:} For multi-application workloads, we generate
100 random mixes and another 100 mixes from the memory intensive subset of
SPEC 2017.  For 4-core workload, 200 Million instructions are used for warm-up
and additional 1 Billion instruction simulated for collecting statistics.
Each CPU keeps executing its workload till the last CPU completes one billion
instructions after warm-up.  For collecting IPC and other data, only the first
billion instructions are considered as the region of interest.

@@

We repeat the same process for 8-core workloads, correspondingly with 16MB
LLC.  The only difference is that 20 million warm-up instructions and 100
million full instructions are executed.  This is done so as to keep the
simulation run-time within reasonable limits as a single 8-core mix takes up
to 3 days to simulate one billion instructions.

\textit{Validation:} We cross-validated our PPF model using SPEC
2006~\cite{SPEC2006} and CloudSuite~\cite{CloudSuite} benchmarks.  For
single-core SPEC 2006, we developed 94 simpoints spread across all the 29
applications. For multi-core, we followed the same methodology as SPEC 2017.
For CloudSuite, we used the traces made available for the 2nd Cache
Replacement Competition (CRC-2)~\cite{CRC_2}.  The traces include 4 4-core
applications with 6 distinct phases per application.

@@

\subsection{Preferchers Simulated}
\label{Method-Prefetchers}
We compared PPF against three of the latest state of the art hardware-only
prefetchers: Best Offset Prefetcher (BOP), DRAM Aware - Access Map Pattern
Matching (DA-AMPM)~\cite{DA_AMPM} and Signature Path Prefecher (SPP).  BOP was
the winner of 2nd Data Prefetching Championship.  DA-AMPM is the enhanced
version of AMPM, modified to account for DRAM row buffer locality.  SPP has
been shown to outperform BOP on SPEC 2006 traces.  For each of these, we
compare their speedups taking no prefetching as the baseline.

\subsection{Developing Features for PPF}
\label{Method-Features}
This section describes the intuition and analysis that went behind developing
the features.  As noted earlier, we developed a set of 9 features that allow
the perceptron throttler to correlate prefetching decision with the program
context.  To study the correlation across each feature, we study statistically
the perceptron weights and try to interpret their distribution.

\textbf{Global Pearson's Correlation}\newline For this experiment, we create a
dump of the state of perceptron weights at the end of all trace execution.
The weights obtained from running all the SPEC 2017 traces are concatenated.
Since the weights are collected at the end of individual trace execution, the
perceptron weights have attained a relatively stable value by now.  Hence,
this dump represents a `snap' of the trained perceptron weights across all the
SPEC 2017 traces.  The intuition here is that the feature with bulk of the
perceptron weights concentrated around 0 or small magnitude numbers show a
weak correlation with the prefetching outcome.  On the other hand, features
with most of the weights saturated around highest value (+15) show a high
positive correlation and the features with weights close to the lowest value
(-16) show a strong negative correlation.

@@

\textbf{Per Trace Correlation} \newline Another important way to look
at the perceptron features is to see how much their contribution
varies across the traces.  Here we give special attention to features
with low P-values in the previous experiment.  Figure XX shows the
variation of P-values three features : \_\_, \_\_ and \_\_; across all
the SPEC 2017 traces.  For simplicity, the traces are arranged in an
increasing order of contribution made by the feature and the trace
names have been omitted.  It can be seen that even features with a low
overall correlation provide useful correlation (magnitude > 0.4) for
XX out of the 95 simpoints developed for SPEC 2017.

\textbf{Trimming Features Using Correlation} \newline Besides
providing interesting insights into prefetching behaviour, P-value can
also be used for feature selection and prefetcher tuning.

Here, we introduce the concept of cross-correlation across features.
Just as we examined correlation of each feature with the final
outcome, we can also study correlation between the features.  We used
the above methodology to eliminate features that provide a little or
no information that has already been captured in other features.

As a part of feature selection, we initially came up a mix of 23
features- primary and composite.  By studying cross correlation of
each of these features against others in a 23x23 matrix, we identified
pairs of features with correlation factor > 0.9 in magnitude and
carefully eliminated redundant features.  Using this approach, we
managed to reduce the feature count to 9.  Thus, in the final
implementation of PPF, no two features have a high correlation
between them.  This way we can be sure that each feature is bringing
in contribution that cannot be captured using other features.

@@

\subsection{Overhead for PPF}
\label{Method-Overheads}

@@

In this section, we analyze the hardware overhead required to
implement PPF.  The Prefetch Filter was enhanced to accommodate
storing of meta-data for perceptron training.  Table
\ref{tab:PF_metadata} depicts the meta-data stored for each entry in
the Prefetch Filter.  Table \ref{tab:PPF_overhead} shows the total
storage overhead of PPF implementation.  The hardware budget for
2nd Data Prefetching championship was 32 KB.  Keeping that in mind 
the considerable speedup PPF obtained over the winner, the extra hardware
budget can be accounted for.  The extra hardware also makes the
overall scheme more scalable than SPP.  In the original SPP paper, it
was demonstrated that adding extra hardware brings little advantage in
terms of performance gain.  The newly added perceptron tables can be
scaled to increase / decrease features depending on the permitted
budget.

### 2018-11-19 results.tex
Mostly editing for English, redundancy, some technical changes, and redundancy.

\section{RESULTS}
\label{Results}

This section discusses the results obtained from running NATCH, in
terms of cache Misses Per Kilo Instructions (MPKI) and Instruction Per
Cycle (IPC) metrics.  For SPEC 2017 suite, first we present the
results for single-threaded workloads then the results for multi-core
workloads are presented.

@@

Figure \_\_ shows the IPC speedup on the memory intensive subset of SPEC 2017. 
All the results have been normalized to baseline \textit{i.e.}, no
prefetching.  As can be seen, NATCH yields a speedup of XX\% over
the baseline.  This is equivalent to XX\% over DA-AMPM, xx\% over BOP
and XX\% over SPP.  Out of the XX benchmarks, NATCH nearly matches
or outperforms most of the prefetchers on XX traces.

At its peak, NATCH manages to get the IPC speedup of a factor of
over \textbf{XX} on the trace \textit{602.gcc\_s-2226B}.  This also
corresponds to speedup gain of \textbf{XX\%} over the next best
prefetcher - SPP.  In general, benchmarks \textit{602.gcc\_s,
  603.bwaves\_s, 605.mcf\_s, 621.wrf\_s} and \textit{649.fotonik3d\_s}
benefit the most from NATCH, as compared to SPP only prefetching.

On the full SPEC 2017 suite, NATCH improves the baseline by XX\%,
which is YY\% more than just SPP.

@@

\subsection{Multi-core Results}
\label{Results-Multi}
In this section, we demonstrate the improvement achieved by NATCH
for a mix of multi-programmed workloads.

\textit{4-core environment} Figure \_\_ shows a comparison of speedups
obtained on 4-core mix of memory intensive subset of SPEC 2017.  We
plot all the 4 prefetchers, normalized to baseline.  The workloads
have been sorted in the increasing order of IPC obtained with running NATCH.
As can be seen,
NATCH offers an IPC improvement by 51.7\% on these traces.  This is
11.9\% over the next best prefetcher SPP; and XX\% and YY\% over BOP
and DA-AMPM respectively.

On a different set of fully random SPEC 2017 4-core mixes, NATCH provides an IPC speedup of 
26.2\% over baseline, which is 5.7\% more than just SPP.

\textit{8-core environment} The sorted comparison of speedup obtained 
on the memory intensive 8-core mixes, is depicted in Fig \_\_.
NATCH improves baseline performance by 38.6\% and that is 10.7\% 
more than using just SPP.

For completely random SPEC 2017 mixes, NATCH improves performance by 
23.6\% over the baseline, corresponding to 4.8\% over SPP.

This increased improvement by NATCH on a multi-core machines is not
surprising.  Since NATCH has an extremely intelligent throttler, it ends up 
eliminating useless prefetches before they can create cache
pollution. This further amplifies the edge that SPP had over BOP in multi-core environment.

@@

\textit{Low-Bandwidth DRAM}
Figure \_\_(a) shows the performance of NATCH against other prefetchers. 
For sake of simplicity, we show 2 traces where NATCH gives the most advantage
over baseline, 2 traces where NATCH doesn't perform as well as the next
best prefetcher (replace with the prefetcher name once results are
available) and GeoMean across memory intensive and across all the 
SPEC 2017 traces. As can be seen, for the memory intensive subset,
 NATCH improves baseline by XX\% and that is XX\% more than using SPP.

\textit{Small LLC}
Figure \_\_(b) shows the similar analysis of NATCH.
For the memory intensive subset, a performance improvement of XX\% 
over baseline is seen with NATCH, XX\% more than just SPP.

\subsection{Cross Validation}
\label{Results-CrossVal}
Here we try to demonstrate the robustness of NATCH by testing it on
completely unseen benchmarks, namely SPEC 2006 and CloudSuite.

Fig \_\_ shows the speed-up achieved by BOP, SPP and NATCH on
selected traces of SPEC 2006, along with the overall geometric mean
over whole of the benchmark and over memory-intensive subset of the
SPEC 2006 benchmark.  As can be seen, NATCH provides a speed-up of
44.4\% over baseline, on the memory intensive subset of SPEC 2006
benchmark.  This corresponds to 6.84\% over SPP and XX\% over BOP.  On
the whole of the SPEC 2006 suite, the speedup obtained is 22.4\%,
which is 3.34\% ahead of SPP.

For 4-core memory intensive mixes, NATCH improves the baseline by 59.2\%, 
8.7\% ahead of SPP. For 8-core memory intensive mixes, the IPC speedup 
over baseline is 47.9\%, 11.5\% ahead of SPP.

>>>>

\section{Results}
\label{Results}

This section discusses the results obtained from running PPF, in
terms of cache Misses Per Kilo Instructions (MPKI) and Instruction Per
Cycle (IPC) metrics.  For SPEC 2017 suite, first we present the
results for single-threaded workloads then the results for multi-core
workloads are presented.

@@

Figure \_\_ shows the IPC speedup on the memory intensive subset of SPEC 2017. 
All the results have been normalized to baseline \textit{i.e.}, no
prefetching.  As can be seen, PPF yields a speedup of XX\% over
the baseline.  This is equivalent to XX\% over DA-AMPM, xx\% over BOP
and XX\% over SPP.  Out of the XX benchmarks, PPF nearly matches
or outperforms most of the prefetchers on XX traces.

At its peak, PPF manages to get the IPC speedup of a factor of
over \textbf{XX} on the trace \textit{602.gcc\_s-2226B}.  This also
corresponds to speedup gain of \textbf{XX\%} over the next best
prefetcher - SPP.  In general, benchmarks \textit{602.gcc\_s,
  603.bwaves\_s, 605.mcf\_s, 621.wrf\_s} and \textit{649.fotonik3d\_s}
benefit the most from PPF, as compared to SPP only prefetching.

On the full SPEC 2017 suite, PPF improves the baseline by XX\%,
which is YY\% more than just SPP.

@@

\subsection{Multi-core Results}
\label{Results-Multi}
In this section, we demonstrate the improvement achieved by PPF
for a mix of multi-programmed workloads.

\textit{4-core environment} Figure \_\_ shows a comparison of speedups
obtained on 4-core mix of memory intensive subset of SPEC 2017.  We
plot all the 4 prefetchers, normalized to baseline.  The workloads
have been sorted in the increasing order of IPC obtained with running PPF.
As can be seen,
PPF offers an IPC improvement by 51.7\% on these traces.  This is
11.9\% over the next best prefetcher SPP; and XX\% and YY\% over BOP
and DA-AMPM respectively.

On a different set of fully random SPEC 2017 4-core mixes, PPF provides an IPC speedup of 
26.2\% over baseline, which is 5.7\% more than just SPP.

\textit{8-core environment} The sorted comparison of speedup obtained 
on the memory intensive 8-core mixes, is depicted in Fig \_\_.
PPF improves baseline performance by 38.6\% and that is 10.7\% 
more than using just SPP.

For completely random SPEC 2017 mixes, PPF improves performance by 
23.6\% over the baseline, corresponding to 4.8\% over SPP.

This increased improvement by PPF on a multi-core machines is not
surprising.  Since PPF has an extremely intelligent throttler, it ends up 
eliminating useless prefetches before they can create cache
pollution. This further amplifies the edge that SPP had over BOP in multi-core environment.

@@

\textit{Low-Bandwidth DRAM}
Figure \_\_(a) shows the performance of PPF against other prefetchers. 
For sake of simplicity, we show 2 traces where PPF gives the most advantage
over baseline, 2 traces where PPF doesn't perform as well as the next
best prefetcher (replace with the prefetcher name once results are
available) and GeoMean across memory intensive and across all the 
SPEC 2017 traces. As can be seen, for the memory intensive subset,
 PPF improves baseline by XX\% and that is XX\% more than using SPP.

\textit{Small LLC}
Figure \_\_(b) shows the similar analysis of PPF.
For the memory intensive subset, a performance improvement of XX\% 
over baseline is seen with PPF, XX\% more than just SPP.

\subsection{Cross Validation}
\label{Results-CrossVal}
Here we try to demonstrate the robustness of PPF by testing it on
completely unseen benchmarks, namely SPEC 2006 and CloudSuite.

Fig \_\_ shows the speed-up achieved by BOP, SPP and PPF on
selected traces of SPEC 2006, along with the overall geometric mean
over whole of the benchmark and over memory-intensive subset of the
SPEC 2006 benchmark.  As can be seen, PPF provides a speed-up of
44.4\% over baseline, on the memory intensive subset of SPEC 2006
benchmark.  This corresponds to 6.84\% over SPP and XX\% over BOP.  On
the whole of the SPEC 2006 suite, the speedup obtained is 22.4\%,
which is 3.34\% ahead of SPP.

For 4-core memory intensive mixes, PPF improves the baseline by 59.2\%, 
8.7\% ahead of SPP. For 8-core memory intensive mixes, the IPC speedup 
over baseline is 47.9\%, 11.5\% ahead of SPP.
